
---
title: "rmap examples---2015-03-08 14:02:54 PDT"
author: ""
output:
  html_document:
    css: ~/Documents/headings.css
    fig_caption: yes
    highlight: default
    number_sections: yes
    theme: readable
    toc: yes
  pdf_document:
    highlight: tango
    number_sections: yes
    pandoc_args: --variable=geometry:margin=0.75in
    toc: yes
fontsize: 12pt
---
<A NAME="top"> </A>
```{r global_options, include=FALSE}
require(knitr)
opts_chunk$set(eval = TRUE, echo = TRUE, fig.height = 10, fig.width = 8, tidy=FALSE)
```
trunk/projects/calibration/07-rmap-v-0.01-06/07-corn-www-rmap/rmap-examples.Rmd

## A `score_statistics` example {#ss-example}
This subsection includes code that is part of a complete `score_statistics` example.  This example continues in Sections [Adding standardized_residuals to our example](#sr-example), [If you don’t have cumulative mortality hazards](#no-mortality-example), and [Adding risk_model_boxplots to our example](#box-plot-example)

```{r eval = TRUE}
library(rmap)
x = read.csv(
  file = "data_set_score_statistics.csv", 
  stringsAsFactors = FALSE)
head(x)
```

```{r eval = TRUE}
e = x$e
t = x$t
risk_model_1 = list(
  r = x$r_1, 
  Lambda_outcome = x$Lambda_outcome_1,
  Lambda_mortality = x$Lambda_mortality_1,
  groupings = list(
    risk = list(k = x$k_1),
    missing = list(K = 4, variable = x$w)))
risk_model_2 = list(
  r = x$r_2, 
  Lambda_outcome = x$Lambda_outcome_2,
  Lambda_mortality = x$Lambda_mortality_2,
  groupings = list(
    risk = list(K = 4),
    missing = list(K = 4, variable = x$w)))
ss = score_statistics(e, t, 
  risk_model_1 = risk_model_1, 
  risk_model_2 = risk_model_2)
ss
```

## Adding `standardized_residuals` to our example

```{r}
sr = standardized_residuals(e, t, 
  risk_model_1 = risk_model_1, 
  risk_model_2 = risk_model_2)
plot(sr, grouping_name = "risk")
```

Prettier...

```{r eval = TRUE}
plot_pars = list(x_max = 0.17,
                 y_max = 9,
                 y_max_ad = 0.33,
                 xlab = "Risk (%)")
plot(sr, grouping_name = "risk", plot_pars)
```

[TOP](#top)

## If you don't have cumulative mortality hazards {#no-mortality-example}

Suppose risk_model_2 has no cumulative mortality hazards.  set `lambda_mortality = NULL` as I have done in the example below.

```{r}
risk_model_2 = list(
  r = x$r_2, 
  Lambda_outcome = x$Lambda_outcome_2,
  Lambda_mortality = NULL,
  groupings = list(
    risk = list(K = 4),
    missing = list(K = 4, variable = x$w)))
ss_no_mortality = score_statistics(e, t, 
  risk_model_1 = risk_model_1, 
  risk_model_2 = risk_model_2)
ss_no_mortality
```

```{r}
sr_no_mortality = standardized_residuals(e, t, 
  risk_model_1 = risk_model_1, 
  risk_model_2 = risk_model_2)
plot(sr_no_mortality, grouping_name = "risk")
```

[TOP](#top)

## Adding `risk_model_boxplots` to our example {#boxplot-example}

We continue with [A score_statistics example](#ss-example)

```{r fig.height = 5}
list_of_risk_models = list(
  risk_model_1 = x$r_1,
  risk_model_2 = x$r_2)
K = 4
risk_quantile_boxplots(list_of_risk_models, K)
```

```{r fig.height = 5}
risk_quantile_boxplots(list_of_risk_models, K, risk_max = 0.6)
```

[TOP](#top)

## A `riskValidate` example {#rv-example}

Below is an example of a complete risk validation grouped analysis of
the sample dataset `x`. 

```{r}
  library(rmap)
  x = read.csv(
    file = "datafRandomSample.csv", 
    stringsAsFactors = FALSE)
  head(x)
```

```{r}
my_e = x$e
my_t = x$t
my_r = x$r
my_design = "randomSample"
my_riskGroup = list(k = x$k)
my_rSummary = "mean"
my_bootstrap = 30

rv = riskValidate(
  e = my_e, t = my_t, r = my_r, design = my_design,
  riskGroup = my_riskGroup, rSummary = my_rSummary,
  bootstrap = my_bootstrap, rvpar = rvparFn())
options(width = 120)
options(digits = 3)
rv
```

[TOP](#top)

## A complete `riskValidateUngrouped` example {#rvu-example}

Below is an example of a complete risk validation ungrouped analysis
using some simulated data.  
```{r}
set.seed(1)
sampleData = df_randomSample_r1_r2(NTotal = 500)
epsilon = nrow(sampleData)^(-1/3)
tStar = 10

rvu = riskValidateUngrouped(
  e = sampleData$e, t = sampleData$t, r = sampleData$r1,
  design = "randomSample", 
  riskGroup = list(
    ungrouped = list(epsilon = epsilon, tStar = tStar)), 
  bootstrap = 20, rvpar = rvparFn(), 
  multicore = FALSE, verbose = FALSE)
rvu
```

[TOP](#top)



## How to call `IAD` to show multiple risk models on the same plot {#iad-example}

The `IAD` function is useful for drawing IADs (Individualized
Attribute Diagrams) for multiple risk models
on the same plot using different colors.

Below, we define the results of two calls to
`riskValidateUngrouped`. The only difference in the two calls is one
uses ` r = sampleData$r1` and the other uses ` r = sampleData$r2`.
```{r}
library(rmap)
set.seed(1)
sampleData = df_randomSample_r1_r2(NTotal = 500)
riskGroup = list(
  ungrouped = list(epsilon = 0.15, tStar = 10))

rvu1b = riskValidateUngrouped(
  e = sampleData$e, t = sampleData$t, r = sampleData$r1,
  design = "randomSample", riskGroup = riskGroup, bootstrap = 20,
  rvpar = FALSE, multicore = FALSE, verbose = FALSE)

rvu2b = riskValidateUngrouped(
  e = sampleData$e, t = sampleData$t, r = sampleData$r2,
  design = "randomSample", riskGroup = riskGroup, bootstrap = 20,
  rvpar = FALSE, multicore = FALSE, verbose = FALSE)

IAD(
  list(rvu1b, rvu2b), 
  rvpar = rvparFn(
    col = c("red", "blue"),
    atX = seq(0, 100, 20), atY = seq(0, 100, 20),
    comment = "20 bootstraps for confidence bands"))
```

[TOP](#top)

## A `caseRiskPercentiles` example {#crp-example}

```{r fig.height = 6, fig.width = 6}
data(data_set_score_statistics)
xxx = data_set_score_statistics
tail(xxx)
cutoff = 0.6
tStar = 10
e = xxx$e
t = xxx$t
r = xxx$r_1
rAnother = xxx$r_2
crp = caseRiskPercentiles(
  cutoff, e, t, tStar, r, rAnother, 
  main = "Case risk percentiles",
  xlab = "Risk model 1", ylab = "Risk model 2",
  cex = 1, pch = 16)
crp
```

```{r fig.height = 6, fig.width = 6}
F1 = ecdf(crp$crp1)
F2 = ecdf(crp$crp2)
plot(F1, verticals = TRUE, do.points = FALSE, col = "red", 
     xlim = c(0, 1),
     main = "Case Risk Percentiles CDF")
plot(F2, verticals = TRUE, do.points = FALSE, col = "blue", add = TRUE)
legend(x = 0, y = 0.95, legend = c("Risk model 1", "Risk model 2"), 
       col = c("red", "blue"),
       lty = c(1, 1))
```

```{r}
crp = caseRiskPercentiles(cutoff, e, t, tStar, r)
crp
```

[TOP](#top)

## A `performanceDifference` example {#pd-example}


```{r}
library(rmap)
set.seed(1)
twoMod = df_twoStage_r1_r2()
twoMod$N
head(twoMod$d)
```

```{r}
pd = performanceDifference(
  e = twoMod$d$e, 
  t = twoMod$d$t,
  rs = list(r1 = twoMod$d$r1, r2 = twoMod$d$r2),
  design = list(c = twoMod$d$c, N = twoMod$N),
  riskGroup = list(K1 = 2, K2 = 2),
  bootstrap = 1000)
pd 
```



# Composite plots {#composite-plots}

```{r}
set.seed(1)
sampleData = df_randomSample_r1_r2(NTotal = 200)

rv1 = riskValidate(
  e = sampleData$e, t = sampleData$t, r = sampleData$r1,
  design = "randomSample",
  riskGroup = list(K = 5), rSummary = "mean",
  bootstrap = 20, rvpar = FALSE)

rv2 = riskValidate(
  e = sampleData$e, t = sampleData$t, r = sampleData$r2,
  design = "randomSample",
  riskGroup = list(K = 5), rSummary = "mean",
  bootstrap = 20, rvpar = FALSE)

rvu1 = riskValidateUngrouped(
  e = sampleData$e, t = sampleData$t, r = sampleData$r1,
  design = "randomSample", 
  riskGroup = list(
    ungrouped = list(epsilon = 0.15, tStar = 10)), 
  bootstrap = 20, rvpar = FALSE, 
  multicore = FALSE, verbose = FALSE)

rvu2 = riskValidateUngrouped(
  e = sampleData$e, t = sampleData$t, r = sampleData$r2,
  design = "randomSample", 
  riskGroup = list(
    ungrouped = list(epsilon = 0.15, tStar = 10)), 
  bootstrap = 20, rvpar = FALSE, 
  multicore = FALSE, verbose = FALSE)

setUpTrellisFn(2, 2, 
               main = "Attribute and Individualized Attribute Diagrams")

attributeDiagramRawFn(pos = c(1, 1), rv1, 
                      col = "#FF0000", lightCol = "#FF666640")
addTextToTrellisFn(pos = c(1, 1), "model 1")

attributeDiagramRawFn(pos = c(1, 2), rv2, 
                      col = "#0000FF", lightCol = "#6666FF40")
addTextToTrellisFn(pos = c(1, 2), "model 2")

IAD_RawFn(pos = c(2, 1), rvu1, 
          col = "#FF0000", lightCol = "#FF666640")
addTextToTrellisFn(pos = c(2, 1), "model 1")

IAD_RawFn(pos = c(2, 2), rvu2, 
          col = "#0000FF", lightCol = "#6666FF40")
addTextToTrellisFn(pos = c(2, 2), "model 2")
```

[TOP](#top)

# Simulation functions {#sim-fns}


```{r}
set.seed(1)
data1 = df_randomSample()
head(data1)
```

```{r}
data2 = df_randomSample_r1_r2()
head(data2)
```

```{r}
data3 = df_twoStage()
data3$N
```

```{r}
head(data3$d)
```

```{r}
data4 = df_twoStage_r1_r2()
data4$N
```

```{r}
head(data4$d)
```

[TOP](#top)

#