---
title: "rmap---v-0.02-03--2017-07-20 14:59:25 PDT"
author: ''
output:
  html_document:
    css: ~/Documents/headings.css
    fig_caption: yes
    highlight: default
    number_sections: yes
    theme: readable
    toc: yes
  pdf_document:
    highlight: tango
    number_sections: yes
    pandoc_args: --variable=geometry:margin=0.75in
    toc: yes
fontsize: 12pt

---
<A NAME="top"> </A>
```{r global_options, include=FALSE}
require(knitr)
opts_chunk$set(eval = TRUE, echo = TRUE, fig.height = 10, fig.width = 8, tidy=FALSE)
```

# Introduction {#introduction}

Personal predictive models use an individual's personal covariates to assign him/her a probability of developing a specific disease within a specified future time period $[0, \text{t_star})$ and before a prespecified competing risk.  Predictive models are often evaluated by comparing their assigned risks to outcome incidences among participants in a cohort study.  Performance of such models can be evaluated using two criteria: calibration and concordance. Model calibration (also called goodness-of-fit) measures how well the model-assigned risks agree
with persons' subsequent observed outcomes. `rmap` offers a grouped goodness-of-fit test and grouped and ungrouped attribute diagrams. Concordance (also called the area under the ROC curve) measures how well a model separates positive and negative outcomes.`rmap` offers estimation and confidence intervals for the concordance as well as ROC  
plots. `rmap` handles random samples, two-stage sample designs, and weighted sample designs.

[TOP](#top)

# Data preparation {#data-preparation}

We assume that you have a data set containing $N$ people, and for the $n$-th person, you have the following data:

## `e`

The event for the subject. Record`e = 1` if the adverse outcome was observed to occur before `t_star` and before the competing event, `e = 2` if the competing risk, such as death from other causes,  was observed to occur before `t_star` and before the adverse outcome, and  `0` if censored or free from adverse outcome and competing risk at time `t_star`.  Acceptable values are `0, 1, 2`.

## `t`

The time until event.  Mathematically speaking, let `t_1` be time until adverse outcome, `t_2` time until competing risk, `t_0` ($\le$ `t_star`) time until censoring . Then `t` is the minimum of `t_1`, `t_2`, `t_0`, `t_star`. And `e` records which of the events occurred first. Acceptable values are `[0, t_star]` (real numbers between `0` and `t_star`, including `0` and `t_star`).

## `r`

The risk assigned by the personal predictive model. This is the probability of adverse outcome occuring before the competing risk and within `[0, t_star)`. Acceptable values are `[0, 1]` (real numbers between `0` and `1`, including `0` and `1`).

## `category` (optional)

If your data were obtained by two-stage sampling, this is the two-stage category indicating which two-stage category the person fell into.   If there were `C` possible categories, acceptable values are `1`, `2`, $\cdots$, `C`. If your data are a cohort sample whose relevant covariates do not match your target population,


[TOP](#top)

# Installing rmap {#installing-rmap}

If you have not yet installed R, download the latest version (at least 3.3.1) for your operating system at:

http://www.r-project.org

Run the R application. To install the rmap package, enter the following lines of code to the R prompt:

```
install.packages("devtools")
library(devtools)
install_github("gailg/rmap")
if("rmap" %in% rownames(installed.packages())){
  print("rmap installed successfully--you are good to go!")
} else {
  print("something went wrong--ask for help")
}
```
If your installation was successful, you should see the message

```
[1] "rmap installed successfully--you are good to go!"
```
[TOP](#top)

# Reading in data {#reading-in-data}

If you saved your data from excel as a CSV (comma separated value) file in your working directory, your next step is to read this CSV file into R.  To practice reading in my example file "data_set_score_statistics.csv", place this file in your working directory and enter the following lines to R. Your result to `head(x)` should look like my output below.


```{r eval = TRUE}
x = read.csv(
    file = "datafRandomSample.csv",
    stringsAsFactors = FALSE)
head(x)
```




```{r}
library(rmap)
set.seed(42)
df1 = df_randomSample(KKK = 4)
x = rmap_random_sample = df1[, c("e", "t", "r", "k")]
head(x)
riskValidate(e = x$e, t = x$t, r = x$r, 
             design = "randomSample",
             riskGroup = list(k = x$k),
                          rSummary = "mean",
                          bootstrap = FALSE, rvpar = rvparFn(col = "darkgreen"))
```

[TOP](#top)


